From noise to knowledge: how randomness generates novel phenomena and reveals information

### 2018-03-02 paper/paper.Rmd
WIP

This pattern, known as 'critical slowing down' for reasons that will become 
obvious, was first proposed an an early warning sign for ecological
transitions by @Wissel1984, but has received considerable attention only more recently,
particularly owing to work in empirical systems [@Scheffer2015].
Our analysis in the origins of noise provides sufficient framework to illustrate
this idea completely.  The premise behind the tipping point just described
is that of a fold bifurucation [which I will not detail here, but see @Holling1973; @Scheffer2009; @Strogatz2001].
The key thing we need to know about this bifurcation is that it involves
changes *derivative* of the deterministic skeleton ($\partial_x \left[ b(x) - d(x) \right]$ in
the notation of our example), which can build up for some time without
significant changes to the location of the equilibrium state $\hat x$.
This derivative did not appear in our marcroscopic equation, $\tfrac{\mathrm{d}x}{\mathrm{d}t}} = b(x) - d(x)$,
but featured regularly in our equations for the noise, such as the 
expected variance, Eq \eqref{var} and autocorrelation, Eq \eqref{autocor}.
This term represents the *stability*  of the equilibrium: how quickly
deviations return to the stationary value.

@@

Noise can provide novel information in other ways as well.  Stochastic
fluctuations act like a continual miniature perturbation experiment,
constantly pushing a system away from an equilibrium to reveal its
behavior elsewhere in state space. Observing the size and
auto-correlation of these fluctuations thus provides insight on the
stability of an equilibrium point, (as can be seen, for instance
in the fluctuation-dissipation result we found for generic birth
death models).  This observation has been exploited with particular
success as a potential *early warning sign* that a system is loosing
stability; possibly due to an approaching catastrophic transition
or saddle-node bifurcation.

This pattern, known as critical slowing
down (in reference to the auto-correlation time in the physics
literature) was first proposed an an early warning sign for ecological
systems by @Wissel1984, but has received considerable empirical as
well as theoretical attention only more recently [@Scheffer2009;
@Scheffer2015].  Increases in two of these early warning indicators,
variance and auto-correlation, are illustrated along with the
timeseries of a simulation of the dynamics of yeast population
growth in Figure \ref{tipping}.  This simulation is based on a mechanistic model
developed by @Dai2012 that closely matches the population dynamics
they observed in serial dilutions of budding yeast _Saccharomyces
cerevisiae_ on a sucrose medium. Because the relevant proteins are
located in the extra-cellular matrix rather than inside the cells,
budding yeast break down sucrose collaboratively, creating an Allee
effect below which the population is too small to sustain itself.
Increasing the serial dilution rate slowly over time drives the
stable population towards a fold bifurcation, which is accompanied
by the patterns of increasing variance and increasing auto-correlation
seen empirically in lab populations [@Dai2012] as well as the
numerical simulation shown in Figure \ref{tipping}.

>>>>

This pattern, known as 'critical slowing down' for reasons that will become 
obvious, was first proposed an an early warning sign for ecological
transitions by @Wissel1984, but has received considerable attention only more recently,
particularly owing to work in empirical systems [@Scheffer2015].
Our analysis in the origins of noise provides sufficient framework to illustrate
this idea completely.  The premise behind the tipping point just described
is that of a fold bifurucation [which I will not detail here, but see @Holling1973; @Scheffer2009; @Strogatz2001].
The key thing we need to know about this bifurcation is that it involves
changes *derivative* of the deterministic skeleton ($\partial_x \left[ b(x) - d(x) \right]_{x=\hat x}$) in
the notation of our example), which can build up for some time without
significant changes to the location of the equilibrium state $\hat x$.
This derivative did not appear in our marcroscopic equation, $\tfrac{\mathrm{d}x}{\mathrm{d}t}} = b(x) - d(x)$,
but featured regularly in our equations for the noise, such as the 
expected variance, Eq \eqref{var} and autocorrelation, Eq \eqref{autocor}.
This derivative represents the *stability*  of the equilibrium: how quickly
deviations return to the stationary value.  
The magnitude of both variance and autocorrelation must increase
as the stability ($\partial_x \left[ b(x) - d(x) \right]_{x=\hat x}$) decreases
prior to a critical tranisition.

@@

Increases in two of these early warning indicators,
variance and auto-correlation, are illustrated along with the
timeseries of a simulation of the dynamics of yeast population
growth in Figure \ref{tipping}.  This simulation is based on a mechanistic model
developed by @Dai2012 that closely matches the population dynamics
they observed in serial dilutions of budding yeast _Saccharomyces
cerevisiae_ on a sucrose medium. Because the relevant proteins are
located in the extra-cellular matrix rather than inside the cells,
budding yeast break down sucrose collaboratively, creating an Allee
effect below which the population is too small to sustain itself.
Increasing the serial dilution rate slowly over time drives the
stable population towards a fold bifurcation, which is accompanied
by the patterns of increasing variance and increasing auto-correlation
seen empirically in lab populations [@Dai2012] as well as the
numerical simulation shown in Figure \ref{tipping}.
